Effect of resonant and random excitation on the LHC beam in the context of pulsed hollow electron lens operation.

### 2018-02-14 resex.tex
comments from Giulio up to 7th turn pulsing section

\section{Experimental and simulation setup\label{sec:exp}}
%put picture on top
\begin{figure*}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		2016 experiment\\
		\includegraphics[width=0.7\linewidth]{bunchfilling_2016.png}	
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		$ $\\
		2017 experiment\\
		\includegraphics[width=0.7\linewidth]{bunchfilling_2017.png}	
	\end{minipage}
	\caption{\label{fig:fill} Filling and excitation scheme for LHC experiments in 2016 (top) and 2017 (bottom). In 2016 the same excitation amplitude was applied on a group of 4 bunch while in 2016 the number was increased to 6 bunches. In 2016 the excitation was also only applied in one plane leading to $2\times24=48$ bunches in total in 2016 and $3\times2\times36=216$ bunches in 2017.}
\end{figure*}
%put picture on top

@@

The transverse feedback is in general active during the whole LHC cycle, therefore it is of very high importance, that the noise introduced by the system does not contribute to any emittance growth. The typical machine cycle requires a short damping time of ~10-20 turns (high feedback gain) for injection oscillation damping, which is increased during the ramp and collisions to ~50-100 turns (lower feedback gain). It has been demonstrated that noise introduced by the transverse feedback system itself is sufficiently low to introduce any measurable emittance growth \cite{adt_noise_emit_2017}. All excitation signals referenced to in this paper, are digitally synthesized in the damper's digital signal processing units and therefore can be considered as ``noise-free''. This point is important as then the effect on the beam can be attributed entirely to the applied excitation and any effect due to unwanted residual noise can be neglected.

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each sub-group of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total. Both schemes are illustrated in Fig.~\ref{fig:fill}.

@@

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The error in 2016 is much larger as no calibration of the ADT was performed beforehand, which was done in 2017. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor during the 2017 experiments and an excitation with $8^{\mathrm{th}}$ turn pulsing. The measured activity is well in line with the expected excitation pattern.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment and an excitation with 8th turn pulsing.}
\end{figure}

\subsection{Overview of simulations\label{sec:sim}}
For the preparation and interpretation of the experiments two different types of simulations were performed:
\begin{enumerate}
	\item tracking of a Gaussian distribution, referred to as ``distribution tracking" in this paper, to obtain loss rates and emittance growth,
	\item Frequency map analysis (FMA) for the visualization of resonances~\cite{fmalaskar}.
\end{enumerate}
For both simulation types, the tracking code Lifetrac \cite{lifetrac} was used. The simulation parameters are summarized in Table~\ref{tab:sim_param}. Further details on the simulations are given in \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
\begin{table}[h]
	\caption{\label{tab:sim_param}%
		Summary of simulation parameters for the distribution tracking and FMA analysis. For details see \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & distribution tracking & FMA \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{450 GeV} \\
			emittance & $3.5~\mu$m& $2.5~\mu$m\\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			particle & 6D Gaussian & equally spaced grid\\
			distribution &  distribution & in $x,y$ up to $10~\sigma$\\
			&  ($10^4$ particles) & with $\frac{\Delta p}{p_0}=0$\\
			turns tracked & $10^6$ turns & $10^4$ turns \\\hline
			optics & \multicolumn{2}{c}{2016 or 2017 injection optics,}\\
			& \multicolumn{2}{c}{$\beta_{\mathrm{IP1/5}}^*=11$~m}\\
			machine  & standard errors & no errors \\
			imperfections &with $a_1=b_1=0$\footnote{Orbit errors are disabled due to different implementation of the $a_1,b_1$ errors in Lifetrac and MAD-X. $b_2$ errors are adjusted to yield an average (over 60 seeds) peak $\beta$-beat of 15\% as expected from optics measurements in the LHC.} &  \\
			octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}}=+19.6~\mathrm{A}$, $I_{\mathrm{MOD}}=-19.6~\mathrm{A}$}\\
			tune $(Q_x,Q_y)$ & \multicolumn{2}{c}{(64.28,59.31) for 2016,}\\
			& \multicolumn{2}{c}{(62.27,60.295) for 2017}\\
			chromaticity & \multicolumn{2}{c}{(+15,+15)}\\
			$(Q'_x,Q'_y)$ & & \\\hline
			transv. aperture & $5.7~\sigma$ & - \\
			long. aperture & $10~\sigma$ & - \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

\section{Experimental and simulation results\label{sec:simex}}
In the following we will jump forth and back between both experimental and simulation results as the experimental results are best understood on the base of the simulations. In summary, the most efficient pulsing patterns could be successfully predicted in simulations, which we will demonstrate on the example of the 2016 experiments (see Sec.~\ref{sec:pattern}). Afterwards we will present the simulation and experimental results of the tested excitation patterns, explicitly the $10^{\mathrm{th}}$~turn, $7^{\mathrm{th}}$~turn and $8^{\mathrm{th}}$~turn pulsing in Secs.~\ref{sec:simex10}--\ref{sec:simex8}) as well as the random excitation in Sec.~\ref{sec:simexran}. As a last chapter (Sec.~\ref{sec:damp}), we will address the influence of the transverse damper in the presence of an external excitation source on emittance, losses and beam distribution.

Quantitative predictions of loss rates and emittance growth proved in general challenging as both observables are influenced in addition by the natural noise present in the LHC, originating for example from vibrations and magnetic field ripple. The observed effect is then the result of the interaction of the external excitation with the natural noise. At injection, this natural noise component is not known and thus presents a very influential but undefined input to the simulations. Beside the natural noise, also collective effects like intra-beam scattering and electron cloud influence the time evolution of emittance and losses. Both are also not taken into account in the simulations, but the effect was minimized during the experiments by choosing an as small as possible bunch intensity of only $0.7\times10^{11}$ protons per bunch.

\subsection{Dependence on the pulsing pattern\label{sec:pattern}}
The effect of each pulsing pattern in comparison to the other patterns can be identified by the means of the loss rate and emittance growth from the distribution tracking. As an example, the simulation results for the 2016 experiment are shown in Fig.~\ref{fig:patternsim} featuring a clear dependence of both parameters on the pattern.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_sigm.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit1.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:patternsim} Relative intensity (top left), bunch length (top right) and horizontal (bottom left) and vertical (bottom right) emittance for different pulsing patterns obtained from distribution tracking based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The resonant excitation is applied in both planes with an amplitude of 96~nrad. No random noise component is added.}
\end{figure}

@@

The experienced reader might have noticed that the emittance in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an increased initial value and then stays almost constant during the entire simulation time, which usually points to an error in the chosen simulation setup. This behavior is however typical for any resonant excitation and can be ascribed to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}, and also any other simulations over $10^6$ turns presented in this paper. Besides an increased emittance, the new distribution also becomes non-Gaussian. This is illustrated by means of the residual in respect to the Gaussian distribution for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing in Fig.~\ref{fig:patternhist}. This initial fast change of the beam distribution is not only an artifact of the simulation, but has been, to the knowledge of the authors, also for the first time experimentally observed during the 2016 and 2017 LHC experiments presented in this paper. The timescale of this fast adjustment is in experiments however much slower than predicted in simulations - on the timescale of minutes compared to seconds. 
\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from distribution tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here $10^4$~turns) minus the initial distribution (here $10^2$~turns), and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure}

In the presence of only a resonant excitation, this distribution is stable in simulations with a constant emittance. Experimentally however, an increase of the emittance growth with the applied excitation amplitude is observed, see for example Figs.~\ref{fig:10thexp},~\ref{fig:7thexp}, not matching the prediction by simulations. By adding a random noise component representative for the natural noise present in the LHC, the constant emittance is changed to an excitation amplitude dependent emittance growth as depicted for example in Figs.~\ref{fig:10thsim},~\ref{fig:7thsim}. An excitation with random uniform noise on the other hand results in constant emittance growth without any initial adjustment phase. The interaction with the natural noise source then only results in an underestimation of loss rates and emittance growth as it is basically equivalent to the application of random noise with an increased amplitude. In addition to the experiments presented in this paper, the effect of a random excitation was also studied in a separate experiment at the LHC for the case of colliding beams \cite{md1433_noise_top_energy,md_noise_bbLHC}.

\subsection{$10^{\mathrm{th}}$ turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. In summary, the following experimental results were obtained:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude (Figs.~\ref{fig:10thexp}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item Change of the beam distribution for the excited bunches (Fig.~\ref{fig:10thexpprof}).	
\end{enumerate}
As loss rates, emittance growth and beam distribution stayed unchanged for the $4$~reference bunches, the above observations can be also truly associated with the effect of the $10^{\mathrm{th}}$~turn pulsing.
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.49\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and relative horizontal (bottom left) and vertical (bottom right) emittance averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments. For all bunches the damper was not active. Relative refers here to the normalization to the initial value. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Fig.~\ref{fig:fill}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation indicated as envelope.}
\end{figure}

Besides the determination of the scaling of the loss rate and emittance growth with the excitation amplitude important for the actual specification of the hollow electron lens, the vertical emittance actually features the strange behavior predicted in simulations: a fast adjustment phase of the beam distribution visible as a rapid increase of the emittance followed by the reach of a new equilibrium distribution apparent by a slower continuous emittance growth. These two phases are indicated in Fig.~\ref{fig:10thexp} in blue and black.

@@

Thanks to the good performance of the Beam Synchrotron Radiation Telescope (BSRT)~\cite{bsrtprofinj}, the distribution changes indirectly observed as changes of the emittance could also be for the first time directly measured. As example Fig.~\ref{fig:10thexpprof} shows the vertical profile of a reference bunch and one bunch experiencing the maximum excitation of $5\cdot\Delta A=A_{\mathrm{max}}$. The distribution of the excited bunch not only changes but becomes clearly non-Gaussian, while the reference bunch stays unchanged.

@@

@@

\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad},

@@

As can be seen from simulations, adding a random noise component not only yields the correct qualitative behavior of the emittance but also increases the loss rates. As the loss rates are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment can be thus further reduced. A large difference however still prevails which is illustrated in Fig.~\ref{fig:10thexploss} by means of a qualitative comparison of the loss rates obtained in experiments and simulations.

@@

		\includegraphics[width=0.8\linewidth]{2016_scale_amp_10v_ran_lbllong_sim.png}

>>>>

\section{Experimental and simulation setup\label{sec:exp}}
%put picture on top
\begin{figure*}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		2016 experiment\\
		\includegraphics[width=0.9\linewidth]{bunchfilling_2016.png}	
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		$ $\\
		2017 experiment\\
		\includegraphics[width=0.9\linewidth]{bunchfilling_2017.png}	
	\end{minipage}
	\caption{\label{fig:fill} Filling and excitation scheme for LHC experiments in 2016 (top) and 2017 (bottom). In 2016 the same excitation amplitude was applied on a group of 4 bunch while in 2016 the number was increased to 6 bunches. In 2016 the excitation was also only applied in one plane leading to $2\times24=48$ bunches in total in 2016 and $3\times2\times36=216$ bunches in 2017.}
\end{figure*}
%put picture on top

@@

The transverse feedback is in general active during the whole LHC cycle, therefore it is critical that the noise introduced by the system does not contribute to any emittance growth. The typical machine cycle requires a short damping time of ~10-20 turns (high feedback gain) for injection oscillation damping, which is increased during the ramp and collisions to ~50-100 turns (lower feedback gain). It has been demonstrated that noise introduced by the transverse feedback system itself is sufficiently low not to introduce any measurable emittance growth \cite{adt_noise_emit_2017}. All excitation signals referenced to in this paper are digitally synthesized in the damper's digital signal processing units and therefore can be considered as ``noise-free''. This point is important as then the effect on the beam can be attributed entirely to the applied excitation and any effect due to unwanted residual noise can be neglected.

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each sub-group of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total and sub-groups of 4 bunches. Both schemes are illustrated in Fig.~\ref{fig:fill}. In order to reduce the error of the measurements, the observables are averaged over each sub-group of bunches experiencing the same excitation amplitude, yielding in total 5~different amplitudes $n\Delta A$ with $n=1,\cdots, 5$ plus the references bunch without excitation for each transverse feedback configuration and excitation plane. Furthermore, the experiments were usually split in 3 different parts:
\begin{description}
	\item[Part 1] No excitation is applied in order to allow the distribution to reach an equilibrium state after injection.
	\item[Part 2] The excitation is applied with a first maximum excitation amplitude of $A_{\mathrm{max,1}}=5\Delta A_1$.
	\item[Part 3] The maximum excitation amplitude is further increased to $A_{\mathrm{max,1}} < A_{\mathrm{max,2}}=5\Delta A_2$.
\end{description}
Each part lasted approximately 10-15~minutes, which is considered to be a long enough time window to allow the beam distribution to reach its new equilibrium state. In the plots of the experimental results shown later in the paper, the maximum excitation amplitude is indicated with black horizontal arrows, while the average of each sub-group of bunches with the same excitation amplitude $n\Delta A$ is distinguished with different colors.

@@

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The error in 2016 is much larger as no calibration of the ADT was performed beforehand, which was done in 2017. The maximum kick strength reachable with the ADT at injection without risking saturation is 96~nrad. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor during the 2017 experiments and an excitation with $8^{\mathrm{th}}$ turn pulsing. The measured activity is well in line with the expected excitation pattern.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment and an excitation with 8th turn pulsing.}
\end{figure}

\subsection{Overview of simulations\label{sec:sim}}
For the preparation and interpretation of the experiments two different types of simulations were performed:
\begin{enumerate}
	\item tracking of a Gaussian distribution, referred to as ``distribution tracking" in this paper, to obtain loss rates and emittance growth,
	\item Frequency map analysis (FMA) for the visualization of resonances~\cite{fmalaskar}.
\end{enumerate}
For both simulation types, the tracking code Lifetrac \cite{lifetrac} was used. The simulation parameters are summarized in Table~\ref{tab:sim_param}. Further details on the simulations are given in Refs.~\cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
\begin{table}[h]
	\caption{\label{tab:sim_param}%
		Summary of simulation parameters for the distribution tracking and FMA analysis. For details see \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & distribution tracking & FMA \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{450 GeV} \\
			emittance & $3.5~\mu$m& $2.5~\mu$m\\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			particle & 6D Gaussian & equally spaced grid\\
			distribution &  distribution & in $x,y$ up to $10~\sigma$\\
			&  ($10^4$ particles) & with $\frac{\Delta p}{p_0}=0$\\
			turns tracked & $10^6$ turns & $10^4$ turns \\\hline
			optics & \multicolumn{2}{c}{2016 or 2017 injection optics,}\\
			& \multicolumn{2}{c}{$\beta_{\mathrm{IP1/5}}^*=11$~m}\\
			machine  & standard errors & no errors \\
			imperfections &with $a_1=b_1=0$\footnote{Orbit errors are disabled due to different implementation of the $a_1,b_1$ errors in Lifetrac and MAD-X. $b_2$ errors are adjusted to yield an average (over 60 seeds) peak $\beta$-beat of 15\% as expected from optics measurements in the LHC.} &  \\
			octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}}=+19.6~\mathrm{A}$, $I_{\mathrm{MOD}}=-19.6~\mathrm{A}$}\\
			tune $(Q_x,Q_y)$ & \multicolumn{2}{c}{(64.28,59.31) for 2016,}\\
			& \multicolumn{2}{c}{(62.27,60.295) for 2017}\\
			chromaticity & \multicolumn{2}{c}{(+15,+15)}\\
			$(Q'_x,Q'_y)$ & & \\\hline
			transv. aperture & $5.7~\sigma$ & - \\
			long. aperture & $10~\sigma$ & - \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

\section{Experimental and simulation results\label{sec:simex}}
In the following we will jump forth and back between both experimental and simulation results as the experimental results are best understood on the base of the simulations. In summary, the most efficient pulsing patterns could be successfully predicted in simulations, which we will demonstrate on the example of the 2016 experiments (see Sec.~\ref{sec:pattern}). Afterwards we will present the simulation and experimental results of the tested excitation patterns, explicitly the $10^{\mathrm{th}}$~turn, $7^{\mathrm{th}}$~turn and $8^{\mathrm{th}}$~turn pulsing in Secs.~\ref{sec:simex10}--\ref{sec:simex8} as well as the random excitation in Sec.~\ref{sec:simexran}. As a last chapter, Sec.~\ref{sec:damp}, we will address the influence of the transverse damper in the presence of an external excitation source on emittance, losses and beam distribution.

Quantitative predictions of loss rates and emittance growth proved in general challenging as both observables are influenced in addition by the natural noise present in the LHC, originating for example from vibrations and magnetic field ripple. The observed effect is then the result of the interaction of the external excitation with the natural noise. At injection, this natural noise component is not known and thus presents a very influential but undefined input to the simulations. Beside the natural noise, also collective effects like intra-beam scattering and electron cloud influence the time evolution of emittance and losses. Both are also not taken into account in these simulations, but the effect was minimized during the experiments by choosing an as small as possible bunch intensity of only $0.7\times10^{11}$ protons per bunch.

\subsection{Dependence on the pulsing pattern\label{sec:pattern}}
The effect of each pulsing pattern in comparison to the other patterns can be identified by the means of the loss rate and emittance growth from the distribution tracking. As an example, the simulation results for the 2016 experiment are shown in Fig.~\ref{fig:patternsim} featuring a clear dependence of both parameters on the pattern.
\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_sigm.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit1.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:patternsim} Relative intensity (top left), bunch length (top right) and horizontal (bottom left) and vertical (bottom right) emittance for different pulsing patterns obtained from distribution tracking based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The resonant and random excitation respectively is applied in both planes with an amplitude of 96~nrad. No random noise component is added in addition.}
\end{figure*}

@@

One may have noticed that the emittance in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an increased initial value and then stays almost constant during the entire simulation time, which usually points to an error in the chosen simulation setup. This behavior is however typical for any resonant excitation and can be ascribed to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}, and also any other simulations over $10^6$ turns presented in this paper. Besides an increased emittance, the new distribution also becomes non-Gaussian. This is illustrated by means of the residual in respect to the Gaussian distribution for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing in Fig.~\ref{fig:patternhist}. This initial fast change of the beam distribution is not an artifact of the simulation, but has been, to the knowledge of the authors, also for the first time experimentally observed during the 2016 and 2017 LHC experiments presented in this paper. The timescale of this fast adjustment is in experiments however much slower than predicted in simulations --- on the timescale of minutes compared to seconds. 
\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from distribution tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here after $10^4$~turns) minus the initial distribution (here after $10^2$~turns) as \%, and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure}

In the presence of only a resonant excitation, this modified distribution is stable in simulations with a constant emittance. As we will see later, an increase of the emittance with the applied excitation amplitude is however observed experimentally not matching the prediction by simulations. By adding a random noise component representative for the natural noise present in the LHC, the constant emittance is changed in simulations to an excitation amplitude dependent emittance growth after an initial adjustment phase. In case of random uniform noise, the excitation leads to constant emittance growth without any initial adjustment phase. The interaction with the natural noise source then only results in an underestimation of loss rates and emittance growth as it is basically equivalent to the application of random noise with an increased amplitude. In addition to the experiments presented in this paper, the effect of a random excitation was also studied in a separate experiment at the LHC for the case of colliding beams \cite{md1433_noise_top_energy,md_noise_bbLHC}.

\subsection{$10^{\mathrm{th}}$ turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. As this was the first time a resonant excitation was tested in the LHC, only in total 48~bunches could be injected in order to still guarantee the protection of the machine. To test different excitation amplitudes during one fill and have enough statistics, the excitation was therefore only applied in one plane. As described earlier, the experiment started with a period of 12~minutes without excitation in order to let the beam distribution fully adjust to its equilibrium state after injection. Afterwards the excitation was applied with the excitation scheme shown in Fig.~\ref{fig:fill} and a maximum excitation amplitude of $A_{\mathrm{max}}=5\Delta A = 48~$nrad for a total of 11~minutes. The excitation was then further increased to the maximum value reachable without saturation of $A_{\mathrm{max}}=5\Delta A = 96~$nrad and kept for 11~minutes. In both cases, the excitation induced the following changes:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude (Figs.~\ref{fig:10thexp}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item Change of the beam distribution for the excited bunches (Fig.~\ref{fig:10thexpprof}).	
\end{enumerate}
As loss rates, emittance growth and beam distribution stayed unchanged for the $4$~reference bunches, the above observations can be also truly associated with the effect of the $10^{\mathrm{th}}$~turn pulsing.
\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and relative horizontal (bottom left) and vertical (bottom right) emittance averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments. For all bunches the damper was not active. Relative refers here to the normalization to the initial value. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad indicated with black arrows, the excitation amplitude was linearly increased for each group of 4~bunches (see Fig.~\ref{fig:fill}). The different colored solid lines labeled with $n\Delta A$ are the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation shown as envelope. The area with a blue background indicates the fast adjustment period of the beam distribution transitioning into a new equilibrium state highlighted with a gray background.}
\end{figure*}
These experimentally obtained scalings of the loss rate and emittance growth with the excitation amplitude can be then used for the actual specification of the hollow electron lens.

Besides the numerical determination of tolerances, the vertical emittance actually features indeed the strange and interesting behavior predicted in simulations: a fast adjustment phase of the beam distribution visible as a rapid increase of the emittance followed by the reach of a new equilibrium distribution apparent by a slower continuous emittance growth. These two phases are indicated in Fig.~\ref{fig:10thexp} in blue and black.

@@

Thanks to the good performance of the Beam Synchrotron Radiation Telescope (BSRT)~\cite{bsrtprofinj}, the distribution changes indirectly observed as changes of the emittance could also be for the first time directly measured. As example Fig.~\ref{fig:10thexpprof} shows the vertical profile of a reference bunch and one bunch experiencing the maximum excitation of $5\cdot\Delta A=A_{\mathrm{max}}=48$~nrad and afterwards 96~nrad. The distribution of the excited bunch not only changes but becomes clearly non-Gaussian, while the reference bunch stays unchanged.
\begin{figure*}[h]

@@

\end{figure*}

@@

\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad}.

@@

As can be seen from simulations, adding a random noise component not only yields the correct qualitative behavior of the emittance but also increases the losses. As the losses are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment can be thus reduced.  To better compare the experimental results for the two different maximum excitation amplitudes applied for slightly different time intervals and also the simulations conducted for a much shorter interval of only 90~s, we define the relative loss rate $R$:
\begin{equation}\label{eqn:lossrate}
R = \frac{I_{\mathrm{start}}-I_{\mathrm{end}}}{I_{\mathrm{start}}\cdot \Delta t}
where $I$ is the intensity and $\Delta t$ the time interval during which the excitation with fixed $A_{\mathrm{max}}$ is applied during the experiments or the duration of the simulation respectively. A qualitative comparison of the loss rates obtained in experiments and simulations is shown in Fig.~\ref{fig:10thexploss} featuring a surprisingly good agreement. It should be also kept in mind that the additional random noise component considerably influences the simulation results and as long as it is unknown should be seen more as a fit parameter than a real experimentally confirmed input parameter. The measured loss rates for $A_{\mathrm{max}}=48$~nrad and $A_{\mathrm{max}}=96$~nrad should furthermore follow the same scaling. Instead the loss rates for the same excitation amplitude are smaller for $A_{\mathrm{max}}=96$~nrad compared to $A_{\mathrm{max}}=48$~nrad. That losses and also emittance growth for a fixed excitation amplitude do not exactly agree for both experiments, is a general tendency observed in all experiments. As the second excitation however starts from an already by the previous excitation changed beam distribution, the difference is however not surprising and can be explained by this difference in initial conditions. For checking the reproducibility of the results in a possible future experiment, it would be interesting to see if the same loss rates are measured if for each $A_{\mathrm{max}}$ a new fill is used. 

@@

%		\includegraphics[width=0.8\linewidth]{2016_scale_amp_10v_ran_lbllong_sim.png}
\includegraphics[width=1.0\linewidth]{2016_scale_amp_10v_ran_lblshort_sim_prstab.png}
